---
title: "B. Analysis of Variance"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{B. Analysis of Variance}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=6, 
  fig.height=4
)
# Legge denne i YAML på toppen for å skrive ut til tex
#output: 
#  pdf_document: 
#    keep_tex: true
# Original:
#  rmarkdown::html_vignette:
#    toc: true
```

```{r setup}
# Start the mixlm R package
library(mixlm, warn.conflicts = FALSE)
```

# Analysis of Variance (ANOVA)
This vignette will focus on univariate ANOVA in various designs including
fixed and mixed effects, and briefly introduce multivariate ANOVA. Models
using random effects will be run through the mixlm package.

The following models will be demonstrated:
* Fixed effect models
  * One-way ANOVA
  * Two-way ANOVA
  * Covariates in ANOVA
  * Fixed effect nested ANOVA
* Linear mixed models
  * Classical LMM
  * Repeated measures LMM

## Simulated data
We will start by simulating some data to use in the examples below. In this
fictitious setup, milk yield is measured as a function of feed type (low/high protein content), cow breed, bull identity,
daughter and age. The three first factors are crossed and balanced, 
while daughter is nested under bull. 

The yield is generated with a linear model with some noise.

```{r}
set.seed(123)
dat <- data.frame(
  feed  = factor(rep(rep(c("low","high"), each=6), 4)),
  breed = factor(rep(c("NRF","Hereford","Angus"), 16)),
  bull  = factor(rep(LETTERS[1:4], each = 12)),
  daughter = factor(c(rep(letters[1:4], 3), rep(letters[5:8], 3), rep(letters[9:12], 3), rep(letters[13:16], 3))),
  age   = round(rnorm(48, mean = 36, sd = 5))
)
dat$yield <- 150*with(dat, 10 + 3 * as.numeric(feed) + as.numeric(breed) + 
                        2 * as.numeric(bull) + 1 * as.numeric(sample(dat$daughter, 48)) + 
                        0.5 * age + rnorm(48, sd = 2))
head(dat)
```

## Fixed effect models
The simplest form of ANOVA is the fixed effect model. This model assumes that
the levels of the factor are fixed and that the only source of variation is
the factor itself.

### One-way ANOVA
Here we assess only the feed effect on yield, i.e., the following model:
$$yield_{an} = \mu + feed_a + \epsilon_{an}$$
where $a$ is the feed level and $n$ is the observation within the feed level.

```{r}
mod <- lm(yield ~ feed, data = dat)
print(anova(mod))
```
In the ANOVA table one can look at Pr(>F) to see if the feed factor has a
significant effect on yield. The summary function can be used to get more
information about the underlying regression model.

```{r}
summary(mod)
```

Basic model assessment can be done using the plot function.
```{r, fig.width=6, fig.height=5}
old.par <- par(mfrow=c(2,1), mar=c(4,4,2,0.5))
plot(mod, which = 1:2, ask=FALSE)
par(old.par)
```

### Two-way crossed effects ANOVA

Here we assess the feed and breed effects and their interaction effect on yield, 
i.e., the following model:
$$yield_{abn} = \mu + feed_a + breed_b + (feed:breed)_{ab} + \epsilon_{abn}$$
where $a$ is the feed level and $n$ is the observation within the feed level.

```{r}
mod <- lm(yield ~ feed*breed, data = dat)
print(anova(mod))
```

If the interaction effect is not significant, we can simplify the model to:
$$yield_{abn} = \mu + feed_a + breed_b + \epsilon_{abn}$$
```{r}
mod <- lm(yield ~ feed+breed, data = dat)
print(anova(mod))
```

### Types of sums of squares
The classical way of defining sums of squares are Type I, Type II, and Type III,
as described in the documentation of the Anova() function in the car package.

```{r}
# Type I - Sequential testing, including one and one effect
print(anova(mod))

# Type II - Testing each term after all others, 
# except ignoring the term's higher-order relatives
print(Anova(mod, type="II"))

# Type III - Testing each term after all others,
# including the term's higher-order relatives
print(Anova(mod, type="III"))
```
For the two-way ANOVA model, the Type I and Type II sums of squares are the same,
while Type III differs. With balanced data, this only happens when the contrast
coding is of the treatment/reference type.

### Contrast codings
The contrast coding can be specified for each factor in the model. The default
is reference coding, but other codings can be specified using the `contrasts`
Since we are running lm() through the mixlm package, we can use the `contrasts`
argument to specify the coding for all effects simultaneously.

```{r}
# Sum-coding, i.e., the sum of all levels is zero and all effects
# are orthogonal in the balanced case.
mod <- lm(yield ~ feed*breed, data = dat, contrasts="contr.sum")
print(Anova(mod, type="III"))

# Weighted coding, i.e., the sum of all levels is zero and the effects
# are weighted by the number of levels, effect-wise.
mod <- lm(yield ~ feed*breed, data = dat, contrasts="contr.weighted")
print(Anova(mod, type="III"))
```

Instead of specifying the contrasts in a specific model, it is also possible to
set the contrasts globally for the session. This means that all subsequent models,
unless specified otherwise, will use the specified contrasts.

```{r}
options(contrasts = c("contr.sum", "contr.poly"))
```

### Covariates in ANOVA
Adding covariates to an ANOVA model is straightforward. Here we add the age
of the cow as a covariate to the two-way ANOVA model. The model becomes:
$$yield_{abn} = \mu + feed_a + breed_b + (feed:breed)_{ab} + age\cdot x_{abn} + \epsilon_{abn},$$
where $x_{abn}$ is the age of the cow and $age$ is its linear coefficient.

```{r}
mod <- lm(yield ~ feed*breed + age, data = dat)
print(Anova(mod, type="II"))
```

### Fixed effect nested ANOVA
In the case of nested factors, we can specify this in the model. In the current
model, we assume that bulls are fixed effects that we are interested in and that
daughters are nested under bulls. In this case, the daughters do not have any
special attributes that would interfere with the estimation of the bull effect,
so we do not have to assume that they are random effects. The model becomes:
$$yield_{abn} = \mu + bull_a + daugter_{b(a)} + \epsilon_{abn},$$

```{r}
mod <- lm(yield ~ bull + daughter%in%bull, data = dat)
print(Anova(mod, type="II"))
```

## Linear mixed models
Adding random effects to a model can be done either using least squares
modelling through the mixlm package or using ML/REML estimation through the
lme4 package (or similar).

### Classical - mixlm
Using the mixlm package, we specify the random effects using the `r()` function.
If we assume that the bull is a random selection from the population of bulls,
we can specify this as a random effect when focusing on feed. The model looks
like a fixed effect model, but the error structure is different:
$$yield_{abn} = \mu + feed_a + breed_b + (feed:breed)_{ab} + \epsilon_{abn}$$

```{r}
mod <- lm(yield ~ feed*r(bull), data = dat)
print(Anova(mod, type="II"))
```
In addition to the ordinary ANOVA table, an overview of variance components
and expected mean squares are printed.

### Restrictions
The mixlm package has unrestricted models as default, but it is possible to
turn on restriction.

```{r}
mod <- lm(yield ~ feed*r(bull), data = dat, unrestricted = FALSE)
print(Anova(mod, type="II"))
```
This effects which tests are performed and how the variance components are
estimated.

### Repeated Measures
A repeated measures model can be a mixed model with a random effect for the
repeated measures, where the repeated measures are nested under the subjects.
Longitudinal data is a common example of repeated measures data, where
the replicates are repetitions over time within subject. If we subset
the simulated data, we can add a longitudinal effect to the model, in this
case a random variation over three time-points. The time effect does not
necessarily need to be random.

```{r}
set.seed(123)
long <- dat[c(1:4,9:12), c("feed", "daughter", "yield")]
long <- rbind(long, long, long)
long$daughter <- factor(long$daughter) # Remove redundant daughter
long$time  <- factor(rep(1:3, each=8))
long$yield <- long$yield + rnorm(24, sd = 100) + rep(c(-200,0,200), each=8)
plot(yield~daughter, data=long)
```

Now we have a feed effect, individuals (daughters), and a time effect repeated inside daughters,
with the model:
$$yield_{ait} = \mu + daughter_i + feed_a + time_t + (feed:time)_{at} + \epsilon_{ait}$$
```{r}
mod <- lm(yield ~ r(daughter) + feed*r(time), data = long, unrestricted=FALSE)
print(Anova(mod, type="II"))
```

### REML
REML estimation can be done directly with the lme4 package, but we can also
do this through the mixlm package, leveraging the r() function.

```{r}
mod <- lm(yield ~ feed*r(bull), data = dat, REML = TRUE)
print(Anova(mod, type="III"))
```
To see how mixlm transforms the model to lme4, we can print the model.

```{r}
print(mod)
```
We observe that (1 | bull) and (1 | feed:bull) are added to the model, which
means that random intercepts are added for both bull and the interaction.

## Multivariate ANOVA (MANOVA)
Basic multivariate ANOVA can be done using the lm() function, if we create
a matrix of responses. In this case, we add a mastitis effect to the model.

```{r}
dat$mastitis <- as.numeric(dat$breed) + as.numeric(dat$feed) + rnorm(48, sd = 1)
```

The model becomes:
$$[yield_{abn} | mastitis_{abn}] = \mu + feed_a + breed_b + (feed:breed)_{ab} + \epsilon_{abn},$$
where each of the model terms now are vectors matching the number of responses.

```{r}
mod <- lm(cbind(yield,mastitis) ~ feed*breed, data = dat)
print(Anova(mod, type="II"))
```

The test statistics are joint for all responses, here in the form of the default Pillai's
test statistics. Other statistics can be produced as follows:

```{r}
print(Anova(mod, type="II", test="Wilks"))
print(Anova(mod, type="II", test="Hotelling-Lawley"))
print(Anova(mod, type="II", test="Roy"))
```

The summary function can be used to get more information
about the underlying regression model, here revealing the regressions are
performed separately for each response.

```{r}
summary(mod)
```
