---
title: "Caldana"
author: "Liland"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(HDANOVA)
library(vegan)
```

# Example wishlist
- HDANOVA: ASCA, APCA, ASCA+,  
- Vegan: PRC,  
- MEDA toolbox: VASCA, GASCA (no joy)

## Comments:
- There are no replicates, so no interactions are used in the models.
- Timepoint 0 is removed to achieve fully crossed design.
- ASCA and ASCA+ should give equal results since the design is balanced?!
- Should we scale the data? Scaling has not been applied yet on ASCA or PRC. VASCA and GASCA apparently autoscale by default.
- For VASCA and GASCA I tried to follow the MEDA toolbox examples and correct them where they were impossible to run (possibly updated functions and old examples?).
- As I read the output of VASCA, when tested on the metabolomics data (including both condition and time), all variables are maximally significant (minimum P-value).
- GASCA on metabolomics data ran for several hours without converging. For the expression data, almost 40 GB of memory was consumed before getting to the main GASCA routine, thus crashing.

```{r cars}
load("caldana_selected.RData")
```

# Remove timepoint 0 to achieve full crossed design
```{r}
# Remove timepoint 0
metabolomics <- metabolomics[metabolomics$time!="0",]
metabolomics$time <- factor(metabolomics$time)
expressions <- expressions[expressions$time!="0",]
expressions$time <- factor(expressions$time)
```

# ASCA(+)
metabolomics ~ condition + time

## Metabolomics
(ASCA)
```{r}
#res <- HDANOVA:::.asca(metabolomics~condition+time, metabolomics)
#scoreplot(res, legendpos = "top")
res2 <- multiblock::asca(metabolomics~condition+time, metabolomics)
expl <- attr(res2$scores$condition, "explvar")

summary(res2)

# Scoreplot
plot(res2$projected$condition[,1]+res2$scores$condition[,1], 
     res2$projected$condition[,2]+res2$scores$condition[,2],
     col=rgb(seq(0,1, length.out=18),seq(1,0, length.out=18),0),
     pch=rep(0:7,each=18),
     xlab = paste0("Comp. 1 (", round(expl[1],1), "%)"),
     ylab = paste0("Comp. 2 (", round(expl[2],1), "%)"),
     main = "Condition"
)
points(res2$scores$condition[,1], res2$scores$condition[,2], pch=rep(0:7,each=18), cex=2)
legend("topright", legend=levels(metabolomics$condition), pch=0:7)
legend("topleft", legend=c("Time 20", "Time 360"), pch=c(0,0), col=rgb(c(0,1),c(1,0),c(0,0)))
```

## Expressions
(ASCA)
```{r}
#res <- HDANOVA:::.asca(metabolomics~condition+time, metabolomics)
#scoreplot(res, legendpos = "top")
res2 <- multiblock::asca(expression~condition+time, expressions)
expl <- attr(res2$scores$condition, "explvar")

summary(res2)

# Scoreplot
plot(res2$projected$condition[,1]+res2$scores$condition[,1], 
     res2$projected$condition[,2]+res2$scores$condition[,2],
     col=rgb(seq(0,1, length.out=18),seq(1,0, length.out=18),0),
     pch=rep(0:7,each=18),
     xlab = paste0("Comp. 1 (", round(expl[1],1), "%)"),
     ylab = paste0("Comp. 2 (", round(expl[2],1), "%)"),
     main = "Condition"
)
points(res2$scores$condition[,1], res2$scores$condition[,2], pch=rep(0:7,each=18), cex=2)
legend("topright", legend=levels(expressions$condition), pch=0:7)
legend("topleft", legend=c("Time 20", "Time 360"), pch=c(0,0), col=rgb(c(0,1),c(1,0),c(0,0)))
```

# PRC
## Metabolomics
(PRC)
```{r}
res_metabolomics <- prc(metabolomics$metabolomics, metabolomics$condition, metabolomics$time)
```

### RDA
```{r}
res_metabolomics
```

### PRC
```{r}
summary(res_metabolomics)
```

### Plot
```{r}
plot(res_metabolomics, species = FALSE, lty=c(1,1,1,1,2,2,3,3), col=c(1,2,3,4,1,3,1,3), legpos = NA, ylim=c(-0.15,0.2))
legend("bottomright",legend = levels(expressions$condition),lty=c(1,1,1,1,2,2,3,3), col=c(1,2,3,4,1,3,1,3), cex=0.7)
```

## Expressions
(PRC)
```{r}
res_expressions <- prc(expressions$expression, expressions$condition, expressions$time)
summary(res_expressions)
```

### RDA
```{r}
res_expressions
```

### PRC
```{r}
summary(res_expressions)
```

### Plot
```{r}
plot(res_expressions, species = FALSE, lty=c(1,1,1,1,2,2,3,3), col=c(1,2,3,4,1,3,1,3), legpos = NA)
legend("topright",legend = levels(expressions$condition),lty=c(1,1,1,1,2,2,3,3), col=c(1,2,3,4,1,3,1,3), cex=0.7)
```
